### 2013-07-16 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished initial draft of nonlinear chapter

For each of the benchmark problems, we will present the results of
computations using FANM and directly compare them to results using a
Newton-Krylov method in order to verify the correctness of FANM and
its applicability to fluid problems. For each case, we will compare
the global minimum, maximum, and average values for fluid velocity,
pressure, and temperature where applicable. In addition, we will
report the tolerance to which the nonlinear residual was converged.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and
\sn{1}{6}. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} and varying temperature
of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
was used as a means of globalization assisting the Newton correction
computation while the forcing term was computed using method 2 from
Shadid's work.

@@

For the verification, at each Rayleigh number variation the global
minimum, maximum, and average values were required for fluid velocity
in the x and y directions, fluid pressure, and fluid
temperature. Tables~\ref{tab:convection_ra1e3_results},
\ref{tab:convection_ra1e4_results},
\ref{tab:convection_ra1e5_results}, and
\ref{tab:convection_ra1e6_results} give the results of these
computations. It should be noted that the negative pressures computed
are due to a relative pressure being reported. For each of the data
points presented, both Newton-Krylov solutions and FANM solutions
computed identical results for all of these global values.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Reynolds numbers
of 100, 300, 500, and
700. Table~\ref{tab:thermal_convection_parameters} gives the physical
parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Reynolds
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} and varying the fluid
velocity in the $x$ direction on the top boundary. In addition to
multigrid preconditioning for the linear solution at each Newton step,
a polynomial line search method was used as a means of globalization
assisting the Newton correction computation as outlined by Pawlowski
in \citep{pawlowski_globalization_2006} while the forcing term was
computed using method 2 from Shadid's work
\citep{shadid_inexact_1997}.

@@

In the literature on this benchmark, solutions at Reynolds numbers of
up to \sn{1}{4} were achieved using a variety of nonlinear solution
techniques. With the FANM method, at Reynolds numbers above 700,
convergence of the linear iterations could not be achieved regardless
of how agressive the multigrid preconditioner was set. In these cases,
the Jacobian operators are ill-conditioned enough that even GMRES
converges slowly relative to the thermal convection cavity
benchmark. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case and others were transport driven by
intertial forces is dominant. Because of this, only solutions to the
lid driven cavity benchmark at Reynolds numbers of up to 700 are
presented.

For the verification, at each Reynolds number variation the global
minimum, maximum, and average values were required for fluid velocity
in the x and y directions, fluid pressure, and fluid
temperature. Tables~\ref{tab:driven_re100_results},
\ref{tab:driven_re300_results}, \ref{tab:driven_re500_results}, and
\ref{tab:driven_re700_results} give the results of these
computations. It should be noted that the negative pressures computed
are due to a relative pressure being reported. For each of the data
points presented, both Newton-Krylov solutions and FANM solutions
computed identical results for all of these global values.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Krylov
method. For each benchmark, we will analyze the iterative performance
of both the Newton solver and the linear solver used to compute the
correction vector at each step. In addition, timing results will be discussed

@@

@@

@@

@@

@@

@@

>>>>

For each of the benchmark problems, we will present the results of
computations using FANM and directly compare them to results using a
Newton-Krylov method in order to verify the correctness of FANM and
its applicability to fluid problems. For each case, we will compare
the global minimum, maximum, and average values for fluid velocity,
pressure, and temperature where applicable. If the values for these
quantities computed using FANM match those computed using a production
Newton-Krylov method, the implementation will be deemed correct for
the purposes of this work. In addition, we will report the tolerance
to which the nonlinear residual was converged as an additional means
of comparison for correctness.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and
\sn{1}{6}. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term was computed at each Newton step
using method 2 from Shadid's work. In this forcing term method,
nonlinear residuals from the two previous Newton steps are used to
inform the choice of the forcing term for the next step.

@@

For the verification calculations, at each Rayleigh number variation
the global minimum, maximum, and average values were recorded for
fluid velocity in the x and y directions, fluid pressure, and fluid
temperature. Tables~\ref{tab:convection_ra1e3_results},
\ref{tab:convection_ra1e4_results},
\ref{tab:convection_ra1e5_results}, and
\ref{tab:convection_ra1e6_results} give the results of these
computations. It should be noted that the negative pressures computed
are due to a relative pressure being reported. For each of the data
points presented, both Newton-Krylov solutions and FANM solutions
computed identical numerical results for all of these global
values. No variation in solution was noted for each of the
computations within the tolerance of the converged nonlinear residual
norms given in
Table~\ref{tab:convection_residual_norm_comparison}. Variations in the
convergence behavior and differences in performance for these
solutions will be analyzed in the following section.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Reynolds numbers
of 100, 300, 500, and
700. Table~\ref{tab:thermal_convection_parameters} gives the physical
parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Reynolds
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} fixed and varying the
fluid velocity in the $x$ direction on the top boundary. In addition
to multigrid preconditioning for the linear solution at each Newton
step, a polynomial line search method was used as a means of
globalization assisting the Newton correction computation as outlined
by Pawlowski in \citep{pawlowski_globalization_2006} while the forcing
term was again computed using method 2 from Shadid's work
\citep{shadid_inexact_1997}.

@@

In the literature on this benchmark, solutions at Reynolds numbers of
up to \sn{1}{4} were achieved using a variety of nonlinear solution
techniques. With the FANM method, at Reynolds numbers above 700
convergence of the linear iterations could not be achieved regardless
of how agressive the multigrid preconditioner was set. In these cases,
the Jacobian operators are ill-conditioned enough that even GMRES
converges slowly relative to the thermal convection cavity
benchmark. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case and others where transport driven
by intertial forces is dominant. Because of this, only solutions to
the lid driven cavity benchmark at Reynolds numbers of up to 700 are
presented.

For the verification calculations, at each Reynolds number variation
the global minimum, maximum, and average values were recorded for the
fluid velocity in the x and y directions and the fluid
pressure. Tables~\ref{tab:driven_re100_results},
\ref{tab:driven_re300_results}, \ref{tab:driven_re500_results}, and
\ref{tab:driven_re700_results} give the results of these
computations. As with the thermal convection problem, the negative
pressures computed are due to a relative pressure being reported. For
each of the data points given, both Newton-Krylov solutions and FANM
solutions computed identical numerical results for all of these global
values. The norm of the converged nonlinear residual for the
calculations is given in
Table~\ref{tab:driven_residual_norm_comparison}. All computed global
values were within these tolerances.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Krylov
method. For each benchmark, we will analyze the iterative performance
of both the Newton solver and the linear solver used to compute the
correction vector at each step. In addition, timing results will be
discussed.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method. Limiting the number of nonlinear iterations required for
solution is critical for efficiency as nonlinear function evaluations,
Jacobian generation, and preconditioner construction are typically
required at every Newton iteration. For all Rayleigh numbers except
\sn{1}{5} the number of Newton iterations required to converge is the
same while the \sn{1}{5} result required an extra iteration. This is
due to the fact that the convergence of the linear residual has a
strong effect on the convergence of the nonlinear residual, much
stronger that the effect on the convergence of the eigenvalue in the
$SP_N$ computations.

As outlined in our discussion on inexact Newton methods, depending on
how converged the linear residual is at each step, it is possible to
throw of the nonlinear iteration and converge on the wrong nonlinear
solution. Although the same preconditioner was used with both the
Newton-Krylov and FANM calculations, the convergence properties of
MCSA and GMRES are different and we therefore expect them to produce
different results at each nonlinear iteration which will in turn
effect the forcing term calculation. If they are different enough, it
is possible that another Newton iteration may be required. Even
considering the extra Newton iteration, the solutions are converged to
effectivly the same residual norm tolerance and produce the same
results as provided by the
verification.

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case. We see that the extra
iteration comes from slightly slower quadratic convergence with FANM
for this particular case. The difference in convergence of the linear
problem comes from the fact that the same multigrid preconditioner
parameters are used for this case as for the \sn{1}{3} and \sn{1}{4}
Rayleigh number cases. As the Rayleigh number increases, the problem
becomes more difficult to solve and therefore a more agressive
preconditioner should maintain iterative performance of the nonlinear
method. This was observed to be the case as for the \sn{1}{6}
Rayleigh number case the number of multigrid cycle applications was
bumped up to 3 from 1 used for the other cases and FANM converged in
the same number of nonlinear iterations as the Newton-Krylov method.

@@

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/ra_1e5_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the \sn{1}{5} Rayleigh number case.} \textit{The
      FANM calculation required an additional nonlinear iteration to
      converge.}}
  \label{fig:ra1e5_convergence}
\end{figure}

In addition to nonlinear iterative performance, performance of the
inner linear solves is also interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear
iterations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solver preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme. Even the
\sn{1}{5} Rayleigh number case that required more Newton iterations
required 5 fewer MCSA iterations to converge than GMRES iterations.

@@

Although the iterative performance of FANM is excellent for the
thermal convection cavity problem, the explicit preconditioning
strategy causes the same issues as was observed for the $SP_N$
equations in the previous
chapter. Table~\ref{tab:convection_speedup_comparison} gives the
speedup of the Newton-Krylov method over FANM. For all Rayleigh
numbers, we see a fairly constant speedup of O(400) when using
Newton-Krylov over FANM meaning that the implementation of the FANM
method used here is significantly slower. As the inexact Newton scheme
in Drekar was modified by simply swapping out the linear solver used
at each Newton step, this increased compute time is coming directly
from using MCSA as the solver and the overhead due to
preconditioning. When preconditioning overhead for the $SP_N$
equations was considered as shown by the data in
Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
approximately two orders of magnitude was noted between the Krylov
methods and MCSA showing consistency between those results and the
fluid benchmark results. In addition, the fairly constant speedup
values show that FANM has effectively the same time complexity as a
Newton-Krylov method and the implementation timings simply differ by a
constant value.

@@

For the lid driven cavity benchmark, performance results were again
compared between the FANM and Newton-Krylov calculations. In this
case, the nonlinear iterative performance results as given by
Table~\ref{tab:driven_nonlinear_iter_comparison} show the same number
of Newton iterations were required to converge the 100, 300, and 500
Reynolds number cases while the FANM calculation with a Reynolds
number of 700 actually converged in 4 fewer iterations. The nonlinear
residual norm as a function of Newton iteration for this case is given
in Figure~\ref{fig:re700_convergence}.

This result can be explained by the fact that as given by
Table~\ref{tab:driven_mcsa_parameters}, although the preconditioner
parameters were the same for both calculations, the MCSA iteration was
pushed harder by doubling the number of histories used for the
computation at each nonlinear iteration. These extra histories
provided better convergence of the linear model at each iteration and
did not result in oversolving, thereby increasing the convergence rate
of the nonlinear model as well.

@@

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/re700_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the 700 Reynolds number case.} \textit{The FANM
      calculation converged in 4 fewer iterations than the
      Newton-Krylov calculation.}}
  \label{fig:re700_convergence}
\end{figure}

The difficulty of this benchmark problem as compared to the thermal
convection cavity problem is readily apparent in the linear iteration
counts as given by Table~\ref{driven_linear_iter_comparison}. As the
opposite of what was observed for that benchmark, lid driven cavity
results show that the linear models were significantly more
ill-conditioned, requiring many more MCSA and GMRES iterations. For
all cases except the 700 Reynolds number case many more MCSA
iterations were required to converge with 64\% as many GMRES
iterations required as MCSA iterations for the 100 Reynolds number
case. For the 700 Reynolds number case, fewer MCSA iterations were
observed due to the fact that 4 fewer nonlinear iterations were
performed. The average number of linear iterations per nonlinear
iterations was still much larger for the FANM calculations.

These iterative results again point out the fundamental flaw of
requiring a spectral radius of less than unity for the MCSA iteration
to converge at each Newton step. As the problem becomes more dominated
by inertial forces, the resulting linear models become increasingly
stiff and difficult to precondition. As a result, convergence requires
many more iterations that need for the flow problem driven by
convection in the previous benchmark if convergence is achieved at
all.

@@

As a final comparison, the timing of both methods was again compared
with the speedup of using the Newton-Krylov solver over FANM reported
in Table~\ref{tab:driven_speedup_comparison}. The speedup values
reported here are of the same magnitude as those reported for the
thermal convection cavity problem and in agreement with those observed
for the $SP_N$ calculations. In all cases, the preconditioning
strategy must be signficantly improved to avoid these large overheads.
